---
title: "Fama-French 5-Factor Model: Calculations & Visualizations"
author: "Group 2"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
required_packages <- c("tidyverse", "tidyquant", "quantmod", "broom", "PerformanceAnalytics", "tinytex")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]

if(length(new_packages)) {
  install.packages(new_packages, repos = "http://cran.us.r-project.org")
}

library(tidyverse)
library(tidyquant)
library(quantmod)
library(broom)
library(PerformanceAnalytics)
```

# Data Ingestion
This section handles the loading of asset data and the Kenneth French factors.

```{r data-ingestion}
# Code to load data from the /data directory
```

# Preprocessing
Merging returns and factor premiums.

```{r preprocessing}
# Data cleaning and merging logic
```

# Descriptive Statistics
Analyzing the 8 selected assets.

```{r descriptive-stats}
# Mean, SD, Skewness, Kurtosis calculations
```

# Regression Modeling
Fitting the 5-factor model for each asset.

```{r model-fitting}
# Multivariate regression execution
```

# Visualizations
Generating high-fidelity plots for the Final Paper.

```{r visualizations}
# ggplot2 visualizations (Rolling beta, Coefficient bars, etc.)
# Export plots to image/paper/ for LaTeX integration
```

# Model Diagnostics
Checking for significance and residuals.

```{r diagnostics}
# Statistical significance and R-squared analysis
```
