---
title: "Fama-French 5-Factor Model: Calculations & Visualizations"
author: "Group 2"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
required_packages <- c("tidyverse", "tidyquant", "quantmod", "broom", "PerformanceAnalytics", "tinytex", "moments", "zoo")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]

if(length(new_packages)) {
  install.packages(new_packages, repos = "http://cran.us.r-project.org")
}

library(tidyverse)
library(tidyquant)
library(quantmod)
library(broom)
library(PerformanceAnalytics)
library(moments)
library(zoo)
```

# Data Ingestion
This section handles the automated collection of asset data from Yahoo Finance and the loading of Kenneth French factors.

```{r data-ingestion}
# 1. Configuration & Tickers
options(download.file.method = "libcurl")

# Using USD-denominated tickers to align with Fama-French USD Factors
# VUSD.L (S&P 500), WSML.L (Small Cap), VDEV.L (Developed), QUAL (Quality), 
# VIG (Dividend), XLK (Info Tech), IAU (Gold), VWRA.L (All-World)
tickers <- c("VUSD.L", "WSML.L", "VDEV.L", "QUAL", "VIG", "XLK", "IAU", "VWRA.L")

if (!dir.exists("data")) {
  dir.create("data")
}

# 2. Automated Data Collection
message("Checking/Updating historical data from Yahoo Finance...")

lapply(tickers, function(symbol) {
  file_path <- paste0("data/", symbol, ".csv")
  
  if (!file.exists(file_path)) {
    tryCatch({
      df <- suppressWarnings(getSymbols(symbol, src = "yahoo", from = "2015-01-01", auto.assign = FALSE))
      if (!is.null(df) && nrow(df) > 0) {
        df_tidy <- as.data.frame(na.omit(df)) %>%
          rownames_to_column(var = "Date") %>%
          as_tibble()
        write_csv(df_tidy, file_path)
        message(paste("Downloaded:", symbol))
      }
    }, error = function(e) message(paste("Failed to download", symbol, ":", e$message)))
  } else {
    message(paste("Using existing local data for:", symbol))
  }
})

# 3. Monthly Return Calculation Function
load_asset_returns <- function(ticker) {
  file_path <- paste0("data/", ticker, ".csv")
  if (!file.exists(file_path)) return(NULL)
  
  df <- read_csv(file_path, show_col_types = FALSE) %>%
    mutate(Date = as.Date(Date)) %>%
    arrange(Date)
  
  adj_col <- grep("Adjusted", names(df), value = TRUE)
  if (length(adj_col) == 0) return(NULL)
  
  xts_df <- xts(df[[adj_col]], order.by = df$Date)
  monthly_rets <- periodReturn(xts_df, period = "monthly", type = "log")
  
  df_ret <- as.data.frame(monthly_rets) %>%
    rownames_to_column(var = "Date") %>%
    mutate(Date = as.Date(Date)) %>%
    rename(!!ticker := monthly.returns)
  
  return(df_ret)
}

# Load assets and merge
asset_data_list <- compact(lapply(tickers, load_asset_returns))

if (length(asset_data_list) == 0) {
  stop("CRITICAL: No asset data found. Check tickers and internet.")
}

asset_returns <- asset_data_list %>% reduce(full_join, by = "Date")
active_tickers <- names(asset_returns)[names(asset_returns) != "Date"]

# 4. Load Kenneth French Factors
ff5_raw <- read_csv("data/F-F_Research_Data_5_Factors_2x3.csv", skip = 3, show_col_types = FALSE)
monthly_end_row <- which(is.na(ff5_raw[[1]]) | grepl("Annual Factors", ff5_raw[[1]]))[1] - 1

ff5_monthly <- ff5_raw[1:monthly_end_row, ] %>%
  rename(Date_Raw = 1) %>%
  mutate(
    Date = as.Date(paste0(Date_Raw, "01"), format = "%Y%m%d"),
    across(c(`Mkt-RF`, SMB, HML, RMW, CMA, RF), ~ as.numeric(.) / 100)
  ) %>%
  select(Date, `Mkt-RF`, SMB, HML, RMW, CMA, RF)

# 5. Temporal Alignment
asset_returns <- asset_returns %>% mutate(Date = floor_date(Date, "month"))
ff5_monthly <- ff5_monthly %>% mutate(Date = floor_date(Date, "month"))
```

# Preprocessing
Merging returns and factor premiums.

```{r preprocessing}
# Merge assets with factors
final_data <- inner_join(asset_returns, ff5_monthly, by = "Date") %>%
  na.omit()

if (nrow(final_data) == 0) {
  stop("ERROR: No overlapping data between assets and factors.")
}

# Calculate excess returns
final_data_long <- final_data %>%
  pivot_longer(cols = all_of(active_tickers), names_to = "Asset", values_to = "Return") %>%
  mutate(Excess_Return = Return - RF)

# 1. Factor Correlation
factor_corr <- final_data %>%
  select(`Mkt-RF`, SMB, HML, RMW, CMA) %>%
  cor()

# 2. Cumulative Returns
cumulative_returns <- asset_returns %>%
  pivot_longer(-Date, names_to = "Asset", values_to = "Return") %>%
  filter(!is.na(Return)) %>%
  group_by(Asset) %>%
  arrange(Date) %>%
  mutate(Cumulative = exp(cumsum(Return))) %>%
  ungroup()

# 3. Summary Stats for Visuals
stats_summary <- final_data_long %>%
  group_by(Asset) %>%
  summarise(
    Mean = mean(Return, na.rm = TRUE),
    SD = sd(Return, na.rm = TRUE),
    .groups = "drop"
  )
```

# Regression Modeling
Fitting the 5-factor model for each asset.

```{r model-fitting}
# Robust regression workflow using nest/map
model_results <- final_data_long %>%
  group_by(Asset) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(Excess_Return ~ `Mkt-RF` + SMB + HML + RMW + CMA, data = .x)),
    tidied = map(model, tidy),
    glanced = map(model, glance)
  ) %>%
  unnest(tidied) %>%
  select(Asset, term, estimate, std.error, statistic, p.value)

# Extract model fitness separately for easier join
model_glance <- final_data_long %>%
  group_by(Asset) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(Excess_Return ~ `Mkt-RF` + SMB + HML + RMW + CMA, data = .x)),
    glanced = map(model, glance)
  ) %>%
  unnest(glanced) %>%
  select(Asset, r.squared, adj.r.squared, p.value)

# Regression Summary Table
regression_summary <- model_results %>%
  select(Asset, term, estimate, p.value) %>%
  pivot_wider(names_from = term, values_from = c(estimate, p.value))

regression_summary
```

# Visualizations
Generating high-fidelity plots for the Final Paper, including new scatter plots and rolling beta analysis.

```{r visualizations}
# 1. Coefficient Bar Chart
coef_plot <- model_results %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate, fill = Asset)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Fama-French 5-Factor Loadings by Asset",
       x = "Risk Factor",
       y = "Beta Coefficient",
       fill = "Asset Ticker") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 2. Model Accuracy (Adjusted R-squared)
rsquared_plot <- model_glance %>%
  ggplot(aes(x = reorder(Asset, adj.r.squared), y = adj.r.squared, fill = adj.r.squared)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  scale_fill_gradient(low = "red", high = "green") +
  labs(title = "Model Explanatory Power (Adjusted R-squared)",
       x = "Asset",
       y = "Adjusted R-squared")

# 3. Cumulative Returns Plot
cum_plot <- cumulative_returns %>%
  ggplot(aes(x = Date, y = Cumulative, color = Asset)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  scale_y_log10() +
  labs(title = "Cumulative Growth of $1 (2015-Present)",
       subtitle = "Logarithmic Scale",
       x = "Year",
       y = "Wealth Index")

# 4. Risk-Return Scatter Plot
# Plotting Realized Mean Return vs. Market Beta
scatter_data <- stats_summary %>%
  select(Asset, Mean, SD) %>%
  left_join(model_results %>% filter(term == "Mkt-RF") %>% select(Asset, estimate), by = "Asset")

scatter_plot <- scatter_data %>%
  ggplot(aes(x = estimate, y = Mean, label = Asset)) +
  geom_point(aes(size = SD, color = Asset)) +
  geom_text(vjust = -1) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Risk-Return Profile: Realized Return vs. Market Beta",
       x = "Market Beta (Loading)",
       y = "Mean Monthly Return",
       size = "Volatility (SD)")

# 5. Rolling Beta Analysis (VUSD.L as example)
# Note: For brevity, we calculate 24-month rolling beta for VUSD.L vs Market
rolling_beta_data <- final_data %>%
  select(Date, VUSD.L, `Mkt-RF`, RF) %>%
  mutate(VUSD_Excess = VUSD.L - RF)

# Function for rolling regression
roll_reg <- function(d) {
  coef(lm(VUSD_Excess ~ `Mkt-RF`, data = as.data.frame(d)))[2]
}

# Calculate rolling beta using zoo
rolling_beta_data$Beta <- zoo::rollapply(rolling_beta_data %>% select(VUSD_Excess, `Mkt-RF`), 
                                         width = 24, FUN = roll_reg, by.column = FALSE, fill = NA)

rolling_beta_plot <- rolling_beta_data %>%
  filter(!is.na(Beta)) %>%
  ggplot(aes(x = Date, y = Beta)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "24-Month Rolling Market Beta: VUSD.L (S&P 500)",
       x = "Date",
       y = "Rolling Beta Loading")

# Create image/paper directory if it doesn't exist
if (!dir.exists("image/paper")) {
  dir.create("image/paper", recursive = TRUE)
}

# Export all plots for LaTeX
ggsave("image/paper/factor_loadings.png", coef_plot, width = 10, height = 6)
ggsave("image/paper/model_accuracy.png", rsquared_plot, width = 8, height = 5)
ggsave("image/paper/cumulative_returns.png", cum_plot, width = 10, height = 6)
ggsave("image/paper/risk_return_scatter.png", scatter_plot, width = 8, height = 6)
ggsave("image/paper/rolling_beta.png", rolling_beta_plot, width = 10, height = 5)

# Display plots
coef_plot
rsquared_plot
cum_plot
scatter_plot
rolling_beta_plot
```

# Model Diagnostics
Checking for significance and model fitness across the portfolio.

```{r diagnostics}
# 1. Model Fitness Summary (R-squared Audit)
fitness_summary <- model_glance %>%
  arrange(desc(adj.r.squared)) %>%
  rename(`R-squared` = r.squared, `Adj R-squared` = adj.r.squared, `Model P-value` = p.value)

# 2. Significance Audit
# Create a summary of which factors are statistically significant (p < 0.05)
significance_summary <- model_results %>%
  mutate(significant = ifelse(p.value < 0.05, "Yes", "No")) %>%
  select(Asset, term, significant) %>%
  pivot_wider(names_from = term, values_from = significant)

# 3. Formatted Coefficient Table (with Significance Stars)
formatted_coeffs <- model_results %>%
  mutate(
    stars = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    estimate_star = paste0(round(estimate, 4), stars)
  ) %>%
  select(Asset, term, estimate_star) %>%
  pivot_wider(names_from = term, values_from = estimate_star)

# Display diagnostic tables
print("--- MODEL FITNESS (R-SQUARED) ---")
print(fitness_summary)

print("--- FACTOR SIGNIFICANCE (P < 0.05) ---")
print(significance_summary)

print("--- COEFFICIENTS WITH SIGNIFICANCE STARS ---")
print(formatted_coeffs)
```
